---
title: "Reproducible Research: Peer Assessment 1"
output:
html_document:
keep_md: true
---


***

## Loading and preprocessing the data

Load library files / set the global options for the document

```{r setoptions}
library(knitr)
library(ggplot2)
library(lubridate)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

Extracting Data, Converting the resultant data frame to tbl_df

```{r load.preprocess.data}
unzip("activity.zip")
dat <- read.csv("activity.csv")
dat <- tbl_df(dat)
```


***

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r mean.steps}
by_day <- group_by(dat, date)
(by_day <- summarize(by_day, total.steps = sum(steps, na.rm = TRUE)))
```

2. Make a histogram of the total number of steps taken each day

```{r histogram}
ggplot(by_day, aes(total.steps)) + 
        geom_histogram(binwidth = 1000, colour = "grey") +
        labs(title = "Histogram of Total Steps taken per day",
             x = "Total Steps Taken",
             y = "Frequency")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r report.mean.median}
(mean.steps <- mean(by_day$total.steps))
(median.steps <- median(by_day$total.steps))
```

Mean number of steps = `r mean.steps`

Median number of steps = `r median.steps`


***


## What is the average daily activity pattern?


1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r by.int}
## Group the data by time intervals
by_int <- group_by(dat, interval)
by_int <- summarise(by_int, average.steps = mean(steps, na.rm = TRUE))
```

```{r}
## Plot Time Series graph, using ggplot
ggplot(by_int, aes(interval, average.steps)) +
        geom_line() +
        labs(title = "Time Series Plot",
             x = "5 Minute Time Intervals",
             y = "Average steps taken (across all days)")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
(max.steps <- paste(by_int[which.max(by_int$average.steps), 1]))
```

Answer: `r max.steps`

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
indices <- is.na(dat$steps)
(total.nas <- sum(indices))
```

Answer: `r total.nas`

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.


Answer: We will be filling missing values in the dataset with the mean for the respective 5-minute interval. We already computed a logical vector `indices` containing indices in the original dataset for which `steps == NA` in the previous question.

*       The tbl_df object `by_int` will come handy to get values to make substitutions to the new data set
*       We can use the logical vector `indices` to get the row numbers corresponding the `steps == NA`
*       Running a `for` loop with the row numbers, we can substitute the NA values in the new dataset.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

Answer: The new dataset (tbl_df object) `newdat` has been computed and printed below:

```{r}
a <- which(indices == TRUE)
newdat <- dat
for (i in a){
        xt <- which(by_int$interval == newdat[i, ]$interval)
        newdat[i, ]$steps <- by_int[xt,]$average.steps
        }
newdat
```



## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels `weekday` and `weekend` indicating whether a given date is a weekday or weekend day.

```{r}
dd <- newdat
dd <- dd %>%
        mutate(day = ymd(paste(date))) %>%
        mutate(day = wday(day, label = FALSE))
dd$factor <- ifelse(dd$day >= 2 & dd$day<= 6, TRUE, FALSE)
dd[which(dd$factor == TRUE),]$factor <- "weekday"
dd[which(dd$factor == FALSE),]$factor <- "weekend"
dd$factor <- as.factor(dd$factor)
dd
table(dd$factor)
```


2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
gdata <- group_by(dd, factor, interval)
pdata <- summarise(gdata, avg = mean(steps))
p <- ggplot(pdata, aes(interval, avg))
p + geom_line() +
        facet_grid(factor ~.) +
        labs(title = "Time series plot",
             x = "5-minute time intervals",
             y = "Average steps taken")
```

